---
title: "BOO 2025"
subtitle: "BMD"
date: "`r Sys.Date()`" 
author: 
 Hala Saab
output:
  html_document:
    code_download: true
    theme: united
    lowlight: tango
    df_print: paged
    toc: true
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---


# Setup

## Clean

As before, we perform several setup steps at the start of the script to ensure our work is reproducible and clear. 

**Download the R markdown template, clean your environment, set variables, and load in `tidyverse`:**

* `root_dir` - project folder
* `low_store` - location of DEG results for the low toxicity compound
* `high_store` - location of DEG results for the high toxicity compound

```{r clean}
# Clean
rm(list=ls())

```

***

## Load data

**Load the DEG results R environment:**

```{r load-data, warning=FALSE, message=FALSE}
metaData  <- read.csv("output/metadata_store.csv")

cpmData  <- read.csv("output/cpm_store.csv")
library(dplyr)
library(tibble)
library(readr)

```

```{r}
cpmData <- cpmData %>% column_to_rownames(var = "X") 


```



```{r}

log2cpm <- as.data.frame(log2(cpmData + 1)) %>%
  rownames_to_column(var = "SampleID")

top_df <- as.data.frame(t(c("Dose", metaData$conc_amt)))
colnames(top_df) <- c("SampleID", metaData$sample_ID)

log2cpm <- rbind(top_df, log2cpm)
log2cpm[1:5, 1:5]
```
```{r}
  low.bmd <- metaData %>%
    filter(compound_ID == "CS1133") %>%
    arrange(conc_ID)
  

  high.bmd <- metaData %>%
    filter(compound_ID == "CS1106") %>%
    arrange(conc_ID)
  
  control.bmd <- metaData %>%
    filter(sample_type == "Control") %>%
    arrange(conc_ID)
  
  controlP1.bmd <- control.bmd %>% ####
  filter(grepl("P1", plate_ID))
  
  controlP6.bmd <- control.bmd %>%
  filter(grepl("P6", plate_ID))
  
  log2cpm.low.bmd <- log2cpm[ , low.bmd$sample_ID]
  rownames(log2cpm.low.bmd) <- c("Dose", rownames(cpmData))
  log2cpm.low.bmd <- log2cpm.low.bmd %>%
    rownames_to_column(var="SampleID")
  
  log2cpm.high.bmd <- log2cpm[ , high.bmd$sample_ID]
  rownames(log2cpm.high.bmd) <- c("Dose", rownames(cpmData))
  log2cpm.high.bmd <- log2cpm.high.bmd %>%
    rownames_to_column(var="SampleID")
 log2cpm.ctrl.bmd <- log2cpm[ , control.bmd$sample_ID]
  rownames(log2cpm.ctrl.bmd) <- c("Dose", rownames(cpmData))
  log2cpm.ctrl.bmd <- log2cpm.ctrl.bmd %>%
    rownames_to_column(var="SampleID")
  
  lowData.bmd <- cbind(log2cpm.low.bmd,log2cpm.ctrl.bmd)
  lowData.bmd[is.na(lowData.bmd)] <- 0
  lowData.bmd<-lowData.bmd[ ,-23]
  
  highData.bmd <- cbind(log2cpm.high.bmd,log2cpm.ctrl.bmd)
  highData.bmd[is.na(highData.bmd)] <- 0
  highData.bmd<-highData.bmd[ ,-23]
  
  
write_tsv(lowData.bmd, "output/log2CPMlow.txt")  
write_tsv(highData.bmd, "output/log2CPMhigh.txt")  
```




# Session Info

```{r session-info}
devtools::session_info()
```

***

